---
title: "Vitamin D Analysis 2"
author: "Carlos Dobler"
date: "`r paste('Last update:', format(Sys.time(), '%B %d %Y (%H:%M)'))`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = 'center')
```


```{r prep, include=FALSE}

library(tidyverse)
library(readxl)
library(patchwork)



# Load altitude and lat data
read_excel("data/vitd_data/state muncip alt and lat.xlsx") %>% 
  slice(-1) %>% 
  rename(State = 1,
         Municip = 2,
         Altitude2 = 3,
         Lat = 4) -> df_latalt
  


# Load 20-49 women data
read_csv("data/vitd_data/Yan database women 2012 2018 over10 with lat and alt.csv") %>% 
  
  # reduce diabetes to single logical col
  rowwise() %>% 
  mutate(Diab = case_when(all(is.na(c_across(Prediab:Diabet2))) ~ NA,
                          sum(c_across(Prediab:Diabet2)) == 0 ~ F,
                          TRUE ~ T
                     )) %>% 
  ungroup() %>% 
  select(-c(Prediab:Diabet2, Year)) %>% 
  
  mutate(Hypert = as.logical(Hypert)) %>%
  
  rename(Lat_zone = 1) %>% 

  # remove outlier
  filter(Age >= 20) %>% 
  
  # join with altitude lat data
  left_join(df_latalt, by = c("State", "Municip")) -> df_20_50



# Load > 60 women
read_excel("data/vitd_data/Women over 60.xlsx") %>%
  filter(SEXO == 2,
         EDAD >= 60) %>% 
  select(-c(1,3,9,10)) %>%
  rename(Age = 1,
         State = 2,
         Municip = 3,
         Lat_zone = 4,
         Altitude = 5,
         vitD = 6) %>%

  group_by(State, Municip) %>%
  mutate(n = n()) %>%
  filter(n >= 10) %>%
  ungroup() %>%
  select(-n) %>% 
  
  # join with altitude and lat data
  left_join(df_latalt, by = c("State", "Municip")) -> df_60
  

# Join both df
df_60 %>% 
  
  # add empty columns
  mutate(BMI = NA,
         Hypert = NA,
         Sugar = NA,
         Diab = NA) %>% 
  
  # reorder cols
  select(names(df_20_50)) %>%
  
  # join
  bind_rows(df_20_50) -> df_20_60


```


## Individual-scale analysis

### Pair-wise correlations
Between continuous variables: Pearson  
Between a binary and a continuous variable: Mann-Whitney  
Between binary variables: Chi-square

#### Women 20-49:

#### Latitude
```{r, fig.width = 10, fig.asp = 0.2}

func_pw2 <- function(df, v){
  
  df %>% 
    select(-c(all_of(v))) %>% 
    names() %>% 
    map(function(c){
      
      # Both numeric = pearson
      if(is.numeric(df %>% pull(v)) & is.numeric(df %>% pull(c))){
        
        t <- cor.test(df %>% pull(v), df %>% pull(c))
        
        d_s <- tibble(estimate = t$estimate %>% round(2),
                      p.value = t$p.value %>% round(2))
        
        ggplot(df, aes(x = .data[[v]], y = .data[[c]])) +
          geom_point(alpha = 0.5) +
          geom_smooth(method = "lm", se = F, color = "red") +
          labs(caption = str_c("r = ", d_s$estimate, "; p = ", d_s$p.value),
               subtitle = c) +
          theme(axis.title = element_blank())
          
      # Numeric + logical = wilcoxon (mann-whitney)
      } else if(is.numeric(df %>% pull(v)) & is.logical(df %>% pull(c))){
        
        t <- wilcox.test(df %>% pull(v) ~ df %>% pull(c))
        
        d_s <- t$p.value %>% round(2)
        
        df %>% 
          filter(!is.na(.data[[v]]), !is.na(.data[[c]])) %>%
          ggplot(aes(x = .data[[v]], y = .data[[c]])) +
          geom_boxplot() +
          labs(caption = str_c("p = ", d_s),
               subtitle = c) +
          theme(axis.title = element_blank())
        
        # Logical + numeric = wilcoxon
      } else if(is.logical(df %>% pull(v)) & is.numeric(df %>% pull(c))){
        
        t <- wilcox.test(df %>% pull(c) ~ df %>% pull(v))
        
        d_s <- t$p.value %>% round(2)
        
        df %>% 
          filter(!is.na(.data[[v]]), !is.na(.data[[c]])) %>%
          ggplot(aes(x = .data[[v]], y = .data[[c]])) +
          geom_boxplot() +
          labs(caption = str_c("p = ", d_s),
               subtitle = c) +
          theme(axis.title = element_blank())
        
      # Logical + logical = chi sq
      } else if(is.logical(df %>% pull(v)) & is.logical(df %>% pull(c))){
        
        t <- chisq.test(df %>% pull(v), df %>% pull(c))
        
        d_s <- t$p.value %>% round(2)
        
        df %>% 
          filter(!is.na(.data[[v]]), !is.na(.data[[c]])) %>% 
          count(.data[[v]], .data[[c]]) %>% 
          ggplot(aes(.data[[v]], .data[[c]])) +
          geom_tile(aes(fill = n), show.legend = F) + 
          geom_label(aes(label = n)) +
          labs(caption = str_c("p = ", d_s),
               subtitle = c) +
          theme(axis.title = element_blank())
        
      }
    }) %>% 
    
    wrap_plots(nrow = 1)
    
} # end of func


func_pw2(df_20_50[,-c(1:3,11)], "Lat")
```

#### Age
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_20_50[,-c(1:3,11)], "Age")
```

#### Altitud
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_20_50[,-c(1:3,11)], "Altitude")
```

#### BMI
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_20_50[,-c(1:3,11)], "BMI")
```

#### Diabetes
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_20_50[,-c(1:3,11)], "Diab")
```

#### Hypertension
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_20_50[,-c(1:3,11)], "Hypert")
```

#### Sugar
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_20_50[,-c(1:3,11)], "Sugar")
```

#### Vitamin D
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_20_50[,-c(1:3,11)], "vitD")
```


#### Women > 60:

#### Age
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_60[,-c(2:4,7)],"Age")
```

#### Altitude
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_60[,-c(2:4,7)],"Altitude")
```

#### Latitude
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_60[,-c(2:4,7)],"Lat")
```

#### VitD
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_60[,-c(2:4,7)],"vitD")
```



### Multiple regression 1
Response variable = vit D  
Explanatory variables = bmi, lat, alt  
Only using data of women 20-49 (no BMI data for > 60)
```{r}

lm(vitD ~ scale(BMI) + scale(Lat) + scale(Altitude), data = df_20_50) %>% summary()
# lm(vitD ~ scale(BMI) + scale(Lat) + scale(Altitude), data = df_20_50) %>% faraway::vif()

```

### Multiple regression 2
Response variable = vit D  
Explanatory variables = lat, alt  
Using all data (ages 20 - > 60)
```{r}

lm(vitD ~ scale(Lat) + scale(Altitude), data = df_20_60) %>% summary()
# lm(vitD ~ scale(Lat) + scale(Altitude), data = df_20_60) %>% faraway::vif()

```




## Municipal-scale analysis
Using only women 20-49 dataset 
```{r, include = FALSE}

read_excel("data/vitd_data/Municipal database 20 to 49_MS&YG version2.xlsx",
           skip = 1) %>% 
  rename_with(~c("State", "Municip", "State_name", "Mun_name", "Deaths", 
                 "Pop", "Deaths_ht", "Alt", "Lat", "Ethnicity", "x")) %>% 
  select(-x) -> df_mun

df_20_50 %>% 
  group_by(State, Municip) %>% 
  summarize(
    sample_size = n(),
    mean_vitD = mean(vitD),
    nmol_30 = sum(vitD < 30)/n(),
    nmol_50 = sum(vitD < 50)/n(),
    nmol_75 = sum(vitD < 75)/n()
  ) %>% 
  
  {right_join(df_mun, ., by = c("State", "Municip"))} -> df_mun

```

### Pair-wise correlations

#### Deaths per 100,000
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun[,-c(1:6,11)], "Deaths_ht")
```

#### Altitude
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun[,-c(1:6,11)], "Alt")
```

#### Latitude
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun[,-c(1:6,11)], "Lat")
```

#### Ethnicity
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun[,-c(1:6,11)], "Ethnicity")
```

#### Mean Vit D
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun[,-c(1:6,11)], "mean_vitD")
```

#### nmol < 30
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun[,-c(1:6,11)], "nmol_30")
```

#### nmol < 50
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun[,-c(1:6,11)], "nmol_50")
```

#### nmol < 75
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun[,-c(1:6,11)], "nmol_75")
```




### Multivariate regression 0

Response variable = vit D  
Explanatory variables = Altitude and Latitude

#### Unscaled variables
The coefficient value (column "Estimate") represents the mean change of deaths/ht given a one-unit shift in the independent variable    

```{r}
lm(mean_vitD ~ Alt + Lat, data = df_mun) %>% summary()
```

#### Scaled variables  
The coefficient value (column "Estimate") represents the mean change of deaths/ht given a one-std-deviation shift in the independent variable (indicates which variable has a bigger effect on deaths/ht).
```{r}
lm(mean_vitD ~ scale(Alt) + scale(Lat), data = df_mun) %>% summary()
```

#### Variance Inflation factor
Confirms collinearity in the model. VIF = 1 is best. VIF > 5 means problematic variable (correlated with other(s))
```{r}
lm(mean_vitD ~ Alt + Lat, data = df_mun) %>% faraway::vif() %>% round(2)
```



### Multivariate regression 1

Response variable = Deaths per 100,000  
Explanatory variables = Altitude and Latitude

#### Unscaled variables
```{r}
lm(Deaths_ht ~ Alt + Lat, data = df_mun) %>% summary()
```

#### Scaled variables  
```{r}
lm(Deaths_ht ~ scale(Alt) + scale(Lat), data = df_mun) %>% summary()
```

#### Variance Inflation factor
```{r}
lm(Deaths_ht ~ Alt + Lat, data = df_mun) %>% faraway::vif() %>% round(2)
```



### Multivariate regression 2

Response variable = Deaths per 100,000  
Explanatory variables = Altitude, Latitude, nmol < 30

#### Unscaled variables
```{r}
lm(Deaths_ht ~ Alt + Lat + nmol_30, data = df_mun) %>% summary()
```

#### Scaled variables
```{r}
lm(Deaths_ht ~ scale(Alt) + scale(Lat) + scale(nmol_30), data = df_mun) %>% summary()
```

#### Variance Inflation factor
```{r}
lm(Deaths_ht ~ Alt + Lat + nmol_30, data = df_mun) %>% faraway::vif() %>% round(2)
```



### Multivariate regression 3

Response variable = Deaths per 100,000  
Explanatory variables = Altitude, Latitude, nmol < 30, ethnicity

#### Unscaled variables
```{r}
lm(Deaths_ht ~ Alt + Lat + nmol_30 + Ethnicity, data = df_mun) %>% summary()
```

#### Scaled variables
```{r}
lm(Deaths_ht ~ scale(Alt) + scale(Lat) + scale(nmol_30) + scale(Ethnicity), data = df_mun) %>% summary()
```

#### Variance Inflation factor
```{r}
lm(Deaths_ht ~ Alt + Lat + nmol_30 + Ethnicity, data = df_mun) %>% faraway::vif() %>% round(2)
```


## Municipal-scale analysis
Using only women > 60 dataset

### Pairwise correlations
#### Ethnicity
```{r, fig.width = 10, fig.asp = 0.2}

read_excel("data/vitd_data/Deaths women over 60 municipal.xlsx",
           skip = 1) %>%
  select(1,2,9) %>% 
  rename_with(~c("State", "Municip", "Deaths_ht")) -> df_mun_60

df_60 %>% 
  group_by(State, Municip) %>% 
  summarize(
    sample_size = n(),
    mean_vitD = mean(vitD),
    nmol_30 = sum(vitD < 30)/n(),
    nmol_50 = sum(vitD < 50)/n(),
    nmol_75 = sum(vitD < 75)/n(),
    
    Alt = max(Altitude),
    Lat = max(Lat)
  ) %>% 
  
  {right_join(df_mun_60, ., by = c("State", "Municip"))} -> df_mun_60

df_mun %>%
  select(State, Ethnicity) %>% 
  {unique(.)} %>% 
  right_join(df_mun_60, by = "State") -> df_mun_60

func_pw2(df_mun_60[,-c(1,3,5)], "Ethnicity")
```

#### Deaths/100,000
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun_60[,-c(1,3,5)], "Deaths_ht")
```

#### Vit D
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun_60[,-c(1,3,5)], "mean_vitD")
```

#### nmol < 30
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun_60[,-c(1,3,5)], "nmol_30")
```

#### nmol < 50
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun_60[,-c(1,3,5)], "nmol_50")
```

#### nmol < 75
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun_60[,-c(1,3,5)], "nmol_75")
```

#### Altitude
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun_60[,-c(1,3,5)], "Alt")
```

#### Latitude
```{r, fig.width = 10, fig.asp = 0.2}
func_pw2(df_mun_60[,-c(1,3,5)], "Lat")
```


### Multivariate regression
Response variable = Deaths per 100,000  
Explanatory variables = Altitude, Latitude

#### Unscaled variables
```{r}
lm(Deaths_ht ~ Alt + Lat, data = df_mun_60) %>% summary()
```

#### Scaled variables
```{r}
lm(Deaths_ht ~ scale(Alt) + scale(Lat), data = df_mun_60) %>% summary()
```
