---
title: "Vitamin D Analysis 2"
author: "Carlos Dobler"
date: "`r paste('Last update:', format(Sys.time(), '%B %d %Y (%H:%M)'))`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = 'center')
```


```{r prep, include=FALSE}

library(tidyverse)
library(readxl)

# Load files
read_csv("data/vitd_data/Yan database women 2012 2018 over10.csv",
         col_types = "iiidddddddii") %>% 
  
  rowwise() %>% 
  mutate(Diab = case_when(all(is.na(c_across(Prediab:Diabet2))) ~ NA_integer_,
                          sum(c_across(Prediab:Diabet2)) == 0 ~ 0L,
                          TRUE ~ 1L
                     )) %>% 
  ungroup() %>% 
  select(-c(Prediab:Diabet2, Year)) -> db_vitd


```


## Case-scale analysis

### Pair-wise correlations

#### Age
```{r, fig.width = 10, fig.asp = 0.2}

func_pw <- function(df, v){
  
  df %>% 
    pivot_longer(cols = -c(State, Municip, {{v}}),
                 names_to = "vars",
                 values_to = "val") -> d
  
  # d %>%
  #   rename(vv = {{v}}) %>% 
  #   nest(data = (-vars)) %>%
  #   mutate(test = map(data, ~ cor.test(.x$vv, .x$val)),
  #          test = map(test, broom::tidy)) %>%
  #   unnest(cols = test) %>%
  #   select(vars, estimate, p.value) -> d_s
  
  d %>%
    rename(vv = {{v}}) %>% 
    filter(!is.na(vv) & !is.na(val)) %>% 
    group_by(vars) %>% 
    summarize(
      estimate = cor.test(vv, val, use = "complete.obs")$estimate %>% round(2),
      p.value = cor.test(vv, val, use = "complete.obs")$p.value %>% round(2)
    ) -> d_s
  
  ggplot(d, aes(x = {{v}}, y = val)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = F, color = "red") +
    facet_wrap(~vars, nrow = 1, scales = "free_y") +
    geom_label(data = d_s, aes(x = -Inf, y = Inf, label = str_c("r = ", round(estimate, 2), "\np = ", round(p.value, 2))),
               hjust = -0.1, vjust = 1.2, alpha = 0.7) +
    theme(axis.title.y = element_blank())
  
}

func_pw(db_vitd, Age)

```

#### Altitud
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(db_vitd, Altitud)
```

#### BMI
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(db_vitd, BMI)
```

#### Diabetes
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(db_vitd, Diab)
```

#### Hypertension
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(db_vitd, Hypert)
```

#### Sugar
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(db_vitd, Sugar)
```

#### Vitamin D
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(db_vitd, vitD)
```


### Multiple regression
Response variable = vit D  
Explanatory variables = age, bmi, hypert, & diab
```{r, include=FALSE}

db_vitd %>% 
  map_df(~sum(is.na(.x))) # count missing obs per var

db_vitd %>%
  select(-c(Sugar, Altitud)) %>% 
  rowwise() %>% 
  mutate(na = case_when(any(is.na(c_across(everything()))) ~ 0,
                        TRUE ~ 1)) %>% 
  ungroup() %>% 
  filter(na == 1) %>% 
  select(-c(State, Municip, na)) %>% # remove: highest num of missing obs
  
  mutate(Hypert = factor(Hypert),
         Diab = factor(Diab)) -> d

d[,-1] %>% 
  mutate(across(everything(), ~as.numeric(.x))) %>% 
  corrr::correlate() %>% 
  corrr::shave()
  

```

```{r}

lm(vitD ~ Age + BMI + Hypert + Diab, data = d) %>% summary()

```
The explanatory variables, together, explain a negligible amount of the variance in Vit D (see R-squared). When controlling for the other variables, both age and BMI have a negative and signficiant relation with Vit D (see column "Pr(>|t|)").


## Municipal-scale analysis
```{r, include = FALSE}

read_excel("data/vitd_data/Municipal database 20 to 49_MS&YG version2.xlsx",
           skip = 1,
           col_names = c("State", "Municip", "State_name", "Mun_name", "Deaths", 
                         "Pop", "Deaths_ht", "Alt", "Lat", "Ethnicity", "x")) %>% 
  select(-x) -> df_mun

db_vitd %>% 
  group_by(State, Municip) %>% 
  summarize(
    sample_size = n(),
    mean_vitD = mean(vitD),
    nmol_30 = sum(vitD < 30)/n(),
    nmol_50 = sum(vitD < 50)/n(),
    nmol_75 = sum(vitD < 75)/n()
  ) %>% 
  
  {right_join(df_mun, ., by = c("State", "Municip"))} -> df_mun

```

### Pair-wise correlations

#### Deaths per 100,000
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(df_mun[,-c(3:6,11)], Deaths_ht)
```

#### Altitude
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(df_mun[,-c(3:6,11)], Alt)
```

#### Latitude
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(df_mun[,-c(3:6,11)], Lat)
```

#### Ethnicity
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(df_mun[,-c(3:6,11)], Ethnicity)
```

#### Mean Vit D
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(df_mun[,-c(3:6,11)], mean_vitD)
```

#### nmol < 30
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(df_mun[,-c(3:6,11)], nmol_30)
```

#### nmol < 50
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(df_mun[,-c(3:6,11)], nmol_50)
```

#### nmol < 75
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(df_mun[,-c(3:6,11)], nmol_75)
```

## Multivariate regression 1
Response variable = mean Vit D  
Explanatory variables = Deaths per 100,000, Altitude, Latitude, & Ethnicity
```{r}

# df_mun[,7:10] %>% 
#   corrr::correlate() %>% 
#   corrr::shave()

lm(mean_vitD ~ Deaths_ht + Alt + Lat + Ethnicity, data = df_mun) %>% summary()

```
Explanatory variables, together, explain 30% of the variance in Vit D at municipal scale. When controlling for other variables, each variable shows a negative relation with vit D, but only deaths per 100,000, altitude, and latitude are significant. If ethnicity is removed from the model, results are almost the same in terms of coefficients, their signs, and R2.

## Multivariate regression 2
Response variable = Deaths per 100,000  
Explanatory variables = mean Vit D, Altitude, Latitude, & Ethnicity
```{r}

lm(Deaths_ht ~ mean_vitD + Alt + Lat + Ethnicity, data = df_mun) %>% summary()

```
Explanatory variables, together, explain around 4% of the variance in Deaths at municipal scale. When controlling for other variables, vit D shows a negative and significant relation with vit D, while latitude shows a positive significant one. Altitude and ethnicity do not seem significant.
