---
title: "Vitamin D Analysis 2"
author: "Carlos Dobler"
date: "`r paste('Last update:', format(Sys.time(), '%B %d %Y (%H:%M)'))`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = 'center')
```


```{r prep, include=FALSE}

library(tidyverse)
# library(readxl)

# Load files
read_csv("data/vitd_data/Yan database women 2012 2018 over10.csv",
         col_types = "iiidddddddii") %>% 
  
  rowwise() %>% 
  mutate(Diab = case_when(all(is.na(c_across(Prediab:Diabet2))) ~ NA_integer_,
                          sum(c_across(Prediab:Diabet2)) == 0 ~ 0L,
                          TRUE ~ 1L
                     )) %>% 
  ungroup() %>% 
  select(-c(Prediab:Diabet2, Year)) -> db_vitd

```


## Pair-wise correlations

### Age
```{r, fig.width = 10, fig.asp = 0.2}

func_pw <- function(v){
  
  db_vitd %>% 
    pivot_longer(cols = -c(State, Municip, {{v}}),
                 names_to = "vars",
                 values_to = "val") -> d
  
  # d %>%
  #   rename(vv = {{v}}) %>% 
  #   nest(data = (-vars)) %>%
  #   mutate(test = map(data, ~ cor.test(.x$vv, .x$val)),
  #          test = map(test, broom::tidy)) %>%
  #   unnest(cols = test) %>%
  #   select(vars, estimate, p.value) -> d_s
  
  d %>%
    rename(vv = {{v}}) %>% 
    filter(!is.na(vv) & !is.na(val)) %>% 
    group_by(vars) %>% 
    summarize(
      estimate = cor.test(vv, val, use = "complete.obs")$estimate %>% round(2),
      p.value = cor.test(vv, val, use = "complete.obs")$p.value %>% round(2)
    ) -> d_s
  
  ggplot(d, aes(x = {{v}}, y = val)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = F, color = "red") +
    facet_wrap(~vars, nrow = 1, scales = "free_y") +
    geom_label(data = d_s, aes(x = -Inf, y = Inf, label = str_c("r = ", round(estimate, 2), "\np = ", round(p.value, 2))),
               hjust = -0.1, vjust = 1.2, alpha = 0.7) +
    theme(axis.title.y = element_blank())
  
}

func_pw(Age)

```

### Altitud
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(Altitud)
```

### BMI
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(BMI)
```

### Diabetes
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(Diab)
```

### Hypertension
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(Hypert)
```

### Sugar
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(Sugar)
```

### Vitamin D
```{r, fig.width = 10, fig.asp = 0.2}
func_pw(vitD)
```


## Multiple regression
Response variable = vit D  
Explanatory variables = age, bmi, hypert, & diab
```{r, include=FALSE}

db_vitd %>% 
  map_df(~sum(is.na(.x)))

db_vitd %>%
  select(-c(Sugar, Altitud)) %>% 
  rowwise() %>% 
  mutate(na = case_when(any(is.na(c_across(everything()))) ~ 0,
                        TRUE ~ 1)) %>% 
  ungroup() %>% 
  filter(na == 1) %>% 
  select(-c(State, Municip, na)) -> d

d[,-1] %>% 
  corrr::correlate() %>% 
  corrr::shave()
  



```

```{r}

lm(vitD ~ Age + BMI + Hypert + Diab, data = d) %>% summary()

```
The explanatory variables, together, explain a negligible amount of the variance in Vit D (see R-squared). When controlling for the other variables, both age and BMI have a negative and signficiant relation with Vit D (see column "Pr(>|t|)").


