---
title: "Vitamin D Analysis 2.2"
author: "Carlos Dobler"
date: "`r paste('Last update:', format(Sys.time(), '%B %d %Y (%H:%M)'))`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = 'center')
```


```{r prep, include=FALSE}

library(tidyverse)
library(readxl)
library(patchwork)

# Load altitude and lat data
read_excel("data/vitd_data/state muncip alt and lat.xlsx") %>% 
  slice(-1) %>% 
  rename(State = 1,
         Municip = 2,
         Altitude2 = 3,
         Lat = 4) -> df_latalt
  


# Load 20-49 women data
read_csv("data/vitd_data/Yan database women 2012 2018 over10 with lat and alt.csv") %>% 
  
  # reduce diabetes to single logical col
  rowwise() %>% 
  mutate(Diab = case_when(all(is.na(c_across(Prediab:Diabet2))) ~ NA,
                          sum(c_across(Prediab:Diabet2)) == 0 ~ F,
                          TRUE ~ T
                     )) %>% 
  ungroup() %>% 
  select(-c(Prediab:Diabet2, Year)) %>% 
  
  mutate(Hypert = as.logical(Hypert)) %>%
  
  rename(Lat_zone = 1) %>% 

  # remove outlier
  filter(Age >= 20) %>% 
  
  # join with altitude lat data
  left_join(df_latalt, by = c("State", "Municip")) -> df_20_50



# Load > 60 women
read_excel("data/vitd_data/Women over 60.xlsx") %>%
  filter(SEXO == 2,
         EDAD >= 60) %>% 
  select(-c(1,3,9,10)) %>%
  rename(Age = 1,
         State = 2,
         Municip = 3,
         Lat_zone = 4,
         Altitude = 5,
         vitD = 6) %>%

  group_by(State, Municip) %>%
  mutate(n = n()) %>%
  filter(n >= 10) %>%
  ungroup() %>%
  select(-n) %>% 
  
  # join with altitude and lat data
  left_join(df_latalt, by = c("State", "Municip")) -> df_60
  

# Join both df
df_60 %>% 
  
  # add empty columns
  mutate(BMI = NA,
         Hypert = NA,
         Sugar = NA,
         Diab = NA) %>% 
  
  # reorder cols
  select(names(df_20_50)) %>%
  
  # join
  bind_rows(df_20_50) -> df_20_60


```


## Individual-scale analysis

### Multivariate regressions
#### SCALED VARIABLES
1.  
Response variable = vit D  
Explanatory variables = lat, alt  
Women 20-49  
```{r}

lm(vitD ~ scale(Lat) + scale(Altitude), data = df_20_50) %>% summary()
lm(vitD ~ scale(Lat) + scale(Altitude), data = df_20_50) %>% faraway::vif()

```


2.  
Response variable = vit D  
Explanatory variables = lat, alt  
Women > 60  
```{r}

lm(vitD ~ scale(Lat) + scale(Altitude), data = df_60) %>% summary()
# lm(vitD ~ scale(Lat) + scale(Altitude), data = df_20_60) %>% faraway::vif()

```

3.  
Response variable = vit D  
Explanatory variables = lat, alt, age  
Women 20-49  
```{r}

lm(vitD ~ scale(Lat) + scale(Altitude) + scale(Age), data = df_20_50) %>% summary()
lm(vitD ~ scale(Lat) + scale(Altitude) + scale(Age), data = df_20_50) %>% faraway::vif()

```

4.  
Response variable = vit D  
Explanatory variables = lat, alt, age  
Women > 60  
```{r}

lm(vitD ~ scale(Lat) + scale(Altitude) + scale(Age), data = df_60) %>% summary()

```

3.  
Response variable = vit D  
Explanatory variables = lat, alt, age  
Women 20-49  
```{r}

lm(vitD ~ scale(Lat) + scale(Altitude) + scale(Age) + scale(BMI), data = df_20_50) %>% summary()
lm(vitD ~ scale(Lat) + scale(Altitude) + scale(Age) + scale(BMI), data = df_20_50) %>% faraway::vif()

```




## Municipal-scale analysis
```{r, include = FALSE}

read_excel("data/vitd_data/Municipal database 20 to 49_MS&YG version2.xlsx",
           skip = 1) %>% 
  rename_with(~c("State", "Municip", "State_name", "Mun_name", "Deaths", 
                 "Pop", "Deaths_ht", "Alt", "Lat", "Ethnicity", "x")) %>% 
  select(-x) -> df_mun

df_20_50 %>% 
  group_by(State, Municip) %>% 
  summarize(
    sample_size = n(),
    mean_vitD = mean(vitD),
    nmol_30 = sum(vitD < 30)/n(),
    nmol_50 = sum(vitD < 50)/n(),
    nmol_75 = sum(vitD < 75)/n(),
    Age = mean(Age),
    BMI = mean(BMI, na.rm = T)
  ) %>% 
  
  {right_join(df_mun, ., by = c("State", "Municip"))} -> df_mun

```

### Multivariate regressions
#### SCALED VARIABLES / WOMEN 20-49

1.  
Response variable = Deaths / 100,000  
Explanatory variables = lat, alt
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt), data = df_mun) %>% summary()
```
VIF  
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt), data = df_mun) %>% faraway::vif() %>% round(2)
```

2.  
Response variable = Deaths / 100,000  
Explanatory variables = lat, alt, vit D
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD), data = df_mun) %>% summary()
```
VIF  
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD), data = df_mun) %>% faraway::vif() %>% round(2)
```

3.  
Response variable = Deaths / 100,000  
Explanatory variables = lat, alt, vit D, nmol < 30
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30), data = df_mun) %>% summary()
```
VIF  
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30), data = df_mun) %>% faraway::vif() %>% round(2)
```

4.  
Response variable = Deaths / 100,000  
Explanatory variables = lat, alt, vit D, nmol < 30, ethnicity
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30) + scale(Ethnicity), data = df_mun) %>% summary()
```
VIF  
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30) + scale(Ethnicity), data = df_mun) %>% faraway::vif() %>% round(2)
```

5.  
Response variable = Deaths / 100,000  
Explanatory variables = lat, alt, vit D, nmol < 30, ethnicity, age
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30) + scale(Ethnicity) + scale(Age), data = df_mun) %>% summary()
```
VIF  
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30) + scale(Ethnicity) + scale(Age), data = df_mun) %>% faraway::vif() %>% round(2)
```

6.  
Response variable = Deaths / 100,000  
Explanatory variables = lat, alt, vit D, nmol < 30, ethnicity, age, bmi
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30) + scale(Ethnicity) + scale(Age) + scale(BMI), data = df_mun) %>% summary()
```
VIF  
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30) + scale(Ethnicity) + scale(Age) + scale(BMI), data = df_mun) %>% faraway::vif() %>% round(2)
```


#### SCALED VARIABLES / WOMEN > 60

```{r}
read_excel("data/vitd_data/Deaths women over 60 municipal.xlsx",
           skip = 1) %>%
  select(1,2,9) %>% 
  rename_with(~c("State", "Municip", "Deaths_ht")) -> df_mun_60

df_60 %>% 
  group_by(State, Municip) %>% 
  summarize(
    sample_size = n(),
    mean_vitD = mean(vitD),
    nmol_30 = sum(vitD < 30)/n(),
    nmol_50 = sum(vitD < 50)/n(),
    nmol_75 = sum(vitD < 75)/n(),
    
    Alt = max(Altitude),
    Lat = max(Lat),
    
    Age = mean(Age)
  ) %>% 
  
  {right_join(df_mun_60, ., by = c("State", "Municip"))} -> df_mun_60

df_mun %>%
  select(State, Ethnicity) %>% 
  {unique(.)} %>% 
  right_join(df_mun_60, by = "State") -> df_mun_60
```


1.  
Response variable = Deaths / 100,000  
Explanatory variables = lat, alt
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt), data = df_mun_60) %>% summary()
```
VIF  
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt), data = df_mun_60) %>% faraway::vif() %>% round(2)
```

2.  
Response variable = Deaths / 100,000  
Explanatory variables = lat, alt, vit D
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD), data = df_mun_60) %>% summary()
```
VIF  
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD), data = df_mun_60) %>% faraway::vif() %>% round(2)
```

3.  
Response variable = Deaths / 100,000  
Explanatory variables = lat, alt, vit D, nmol < 30
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30), data = df_mun_60) %>% summary()
```
VIF  
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30), data = df_mun_60) %>% faraway::vif() %>% round(2)
```

4.  
Response variable = Deaths / 100,000  
Explanatory variables = lat, alt, vit D, nmol < 30, ethnicity
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30) + scale(Ethnicity), data = df_mun_60) %>% summary()
```
VIF  
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30) + scale(Ethnicity), data = df_mun_60) %>% faraway::vif() %>% round(2)
```

5.  
Response variable = Deaths / 100,000  
Explanatory variables = lat, alt, vit D, nmol < 30, ethnicity, age
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30) + scale(Ethnicity) + scale(Age), data = df_mun_60) %>% summary()
```
VIF  
```{r}
lm(Deaths_ht ~ scale(Lat) + scale(Alt) + scale(mean_vitD) + scale(nmol_30) + scale(Ethnicity)  + scale(Age), data = df_mun_60) %>% faraway::vif() %>% round(2)
```

